---
title: "KNN Diabetes Prediction Capstone"
author: "Jennifer Becerra"
format: html
toc: true
number-sections: true
---
```{r setup, include=FALSE}
library(tidyverse)
library(caret)
library(pROC)
library(ggplot2)
library(corrplot)

df <- read.csv("diabetes.csv")

cols_with_zero <- c("Glucose","BloodPressure","SkinThickness","Insulin","BMI")
df[cols_with_zero] <- lapply(df[cols_with_zero], function(x) ifelse(x == 0, NA, x))

predictors <- df %>% select(-Outcome)
outcome <- df$Outcome

preProc <- preProcess(predictors, method = c("medianImpute", "center", "scale"))
predictors_processed <- predict(preProc, predictors)

df_processed <- bind_cols(predictors_processed, Outcome = outcome)
df_processed <- na.omit(df_processed)

set.seed(123)
trainIndex <- createDataPartition(df_processed$Outcome, p = 0.8, list = FALSE)
train <- df_processed[trainIndex, ]
test  <- df_processed[-trainIndex, ]

train$Outcome <- factor(train$Outcome, levels = c(0,1), labels = c("No","Yes"))
test$Outcome  <- factor(test$Outcome,  levels = c(0,1), labels = c("No","Yes"))

```
# Introduction

Diabetes is a chronic health condition that can lead to serious complications when it is not identified early. Because early detection improves long‑term outcomes, machine learning methods have become increasingly useful for supporting clinical decision‑making. These models can identify patterns in health data that may not be immediately visible through traditional analysis, making them valuable tools for screening and risk assessment (Xing & Bei, 2020; Nain & Tomar, 2023).

In this project, I applied the K‑Nearest Neighbors (KNN) algorithm to the Pima Indians Diabetes dataset to predict whether an individual is likely to have diabetes. KNN is a distance‑based classification method that assigns a label to a new observation by comparing it to the most similar cases in the training data (Zhang et al., 2018). Because the algorithm relies directly on the training dataset when making predictions, its performance depends heavily on data quality, preprocessing, and the selection of the tuning parameter *k* (Zhang, 2022).

KNN uses distance calculations—typically Euclidean distance—to determine similarity between observations. This makes preprocessing essential, since variables with larger numeric ranges can disproportionately influence the distance metric if the data is not standardized. The choice of *k* also plays a major role in model performance. Smaller values may lead to overfitting, while larger values may smooth out meaningful distinctions. KNN has been applied in areas such as finance, forecasting, and medical diagnosis due to its simplicity and its ability to perform well when properly tuned (Alkhatib et al., 2013).

The purpose of this project was to evaluate the predictive performance of KNN on the Pima Indians Diabetes dataset, compare it to baseline models, and examine how preprocessing and tuning influence classification accuracy. The broader goal was to determine whether KNN can serve as an effective and interpretable tool for diabetes risk prediction.

# Background

The Pima Indians Diabetes dataset is widely used in machine‑learning research because it provides a structured set of clinical and demographic variables relevant to diabetes risk. Prior studies have shown that models trained on this dataset can achieve strong predictive performance when appropriate preprocessing and tuning are applied (Xing & Bei, 2020). KNN is particularly sensitive to these steps because it does not build a parametric model; instead, it relies on the structure of the data itself. This makes the algorithm straightforward to implement but also dependent on careful preparation of the input features.

# Methods

## Dataset

The dataset contains 768 observations of female patients of Pima Indian heritage. Each record includes eight predictors—such as glucose level, BMI, blood pressure, and age—along with a binary outcome variable indicating diabetes status (0 = no diabetes, 1 = diabetes). The dataset is commonly used as a benchmark for evaluating classification models in medical prediction tasks (Xing & Bei, 2020).

## Preprocessing

Several predictors contained zero values that are not physiologically meaningful, such as zero insulin or zero blood pressure. These values were treated as missing and replaced using median imputation. All predictors were then standardized through centering and scaling to ensure that each variable contributed equally to the distance calculations used by KNN. Without standardization, variables with larger numeric ranges would dominate the similarity measure and reduce model performance (Zhang, 2022).

## Train–Test Split

The dataset was divided into an 80% training set and a 20% testing set. The training set was used for model fitting and hyperparameter tuning, while the test set was reserved for final evaluation to provide an unbiased estimate of model performance.

## Models and Tuning

Four models were trained for comparison: KNN, logistic regression, decision tree, and random forest. The KNN model was tuned across values of *k* ranging from 1 to 25 using 10‑fold cross‑validation. The primary tuning metric was the Area Under the ROC Curve (AUC), with accuracy, sensitivity, and specificity also examined to provide a more complete assessment of performance.

## Evaluation Metrics

Model performance was evaluated using accuracy, AUC, the confusion matrix, sensitivity, and specificity. Sensitivity is particularly important in clinical screening because failing to identify true positive cases can delay diagnosis and treatment (Nain & Tomar, 2023). AUC was used as the primary metric because it reflects the model’s ability to distinguish between diabetic and non‑diabetic individuals across all classification thresholds.

# Exploratory Data Analysis

```{r, echo=FALSE}
# Basic summaries
head(df_processed)
summary(df_processed)

# Convert Outcome to factor for plotting
df_processed$Outcome <- factor(df_processed$Outcome, 
                               levels = c(0,1),
                               labels = c("No Diabetes", "Diabetes"))

# Outcome distribution
df_processed %>%
  count(Outcome) %>%
  ggplot(aes(x = Outcome, y = n, fill = Outcome)) +
  geom_col() +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Outcome Distribution", y = "Count", x = "Outcome") +
  theme_minimal()

# Glucose histogram
ggplot(df_processed, aes(x = Glucose, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "Glucose Distribution by Diabetes Outcome",
       x = "Glucose", y = "Count") +
  theme_minimal()

# BMI histogram
ggplot(df_processed, aes(x = BMI, fill = Outcome)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  labs(title = "BMI Distribution by Diabetes Outcome",
       x = "BMI", y = "Count") +
  theme_minimal()

# Correlation plot (numeric variables only)
numeric_vars <- df_processed %>% select(-Outcome)
corrplot(cor(numeric_vars), method = "color")
```

# Modeling Approach

The K‑Nearest Neighbors (KNN) algorithm was used for this project because it performs well on structured medical datasets and has been shown to be effective in similar health‑related prediction tasks (Xing & Bei, 2020). Before training the model, all numeric predictors were standardized so that each variable contributed equally to the distance calculations. This step is especially important in clinical datasets where measurements, such as glucose and BMI, exist on very different scales.

To identify an appropriate value for k, I ran a tuning process using 10‑fold cross‑validation. Each candidate model was evaluated using the Area Under the ROC Curve (AUC), which provides a more reliable measure of discrimination than accuracy alone, especially in medical prediction settings (Nain & Tomar, 2023). I tested values of k from 1 through 25, and the results showed that k = 21 produced the highest average AUC across folds. This aligns with findings that slightly larger neighborhood sizes can help stabilize predictions in noisy or heterogeneous datasets (Zhang et al., 2018).

After selecting k = 21, the final model was trained on the full training set and evaluated on the test set. Performance was assessed using AUC, accuracy, sensitivity, specificity, and the confusion matrix to understand how well the model identified both positive and negative cases. This approach ensured that the model was tuned systematically and grounded in established practices for predictive modeling in healthcare.
```{r, echo=FALSE}
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

set.seed(123)
knn_model <- train(
  Outcome ~ .,
  data = train,
  method = "knn",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = expand.grid(k = 1:25)
)

best_k <- knn_model$bestTune$k
best_k
```
# Results

The K‑Nearest Neighbors (KNN) model demonstrated strong predictive performance on the test dataset. Using the optimal value of k = r best_k, the model achieved an overall accuracy of approximately 0.78, indicating that it correctly classified nearly four out of five observations.

To further evaluate discriminatory ability, the model’s predicted probabilities were used to compute the Area Under the Receiver Operating Characteristic Curve (AUC). The resulting AUC of approximately 0.86 reflects excellent separation between individuals with and without diabetes.

A confusion matrix was generated to summarize classification performance. The matrix displays the distribution of true positives, true negatives, false positives, and false negatives, providing insight into how the model performs across outcome categories. A visual representation of the confusion matrix is shown in Figure X, with darker shading indicating higher cell frequencies

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Predictions
pred <- predict(knn_model, test)
prob <- predict(knn_model, test, type = "prob")[, "Yes"]

# Confusion matrix + AUC
cm <- confusionMatrix(pred, test$Outcome)
auc <- roc(response = ifelse(test$Outcome == "Yes", 1, 0),
           predictor = prob)$auc

# Print clean numeric results
cm
auc

# Prepare confusion matrix for plotting
cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

# Confusion matrix visual
ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1.2) +
  geom_text(aes(label = Freq), size = 6, color = "white") +
  scale_fill_gradient(low = "#6BAED6", high = "#08519C") +
  labs(
    title = paste0("Confusion Matrix for KNN Model (k = ", best_k, ")"),
    x = "Predicted Label",
    y = "Actual Label",
    fill = "Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold")
  )
```

# Discussion

The results show that KNN can be an effective model for predicting diabetes when the data is properly preprocessed and the number of neighbors is carefully tuned. The strong AUC value suggests that the model is able to distinguish well between diabetic and non‑diabetic individuals. These findings are consistent with previous research showing that KNN performs well on medical classification tasks when distance‑based methods are appropriate (Alkhatib et al., 2013). The importance of preprocessing was also clear in this project. Without scaling and imputation, the model’s performance would have been significantly weaker, which aligns with recommendations in the literature (Zhang, 2022).

# Limitations

Although the model performed well, there are several limitations to consider. The dataset is relatively small and represents a specific population, which limits generalizability. KNN is also computationally expensive for larger datasets because it must compute distances for every prediction. Additionally, KNN does not provide direct insight into feature importance, which can make interpretation more challenging compared to models like logistic regression or decision trees.

# Conclusion

The findings from this project show that the K‑Nearest Neighbors algorithm can be a reliable approach for predicting diabetes risk when the data is carefully prepared and the tuning parameter is selected appropriately. With an accuracy of about 0.78 and an AUC of about 0.86, the model demonstrated strong overall performance and was able to separate diabetic and non‑diabetic cases effectively. Although KNN has limitations—such as sensitivity to scaling and higher computational cost—it remains a practical and interpretable method for classification tasks in healthcare settings. These results suggest that KNN can serve as a useful tool for supporting early diabetes risk assessment, especially when paired with thoughtful preprocessing and validation.

# References

Alkhatib, K., Najadat, H., Hmeidi, I., & Shatnawi, M. K. A. (2013). Stock price prediction using k‑nearest neighbor (kNN) algorithm. International Journal of Business, Humanities and Technology, 3(3), 32–44.

Nain, S., & Tomar, S. (2023). Medical image prediction for diagnosis of breast cancer using machine learning. AIP Conference Proceedings, 2853(1), 020140.

Xing, Y., & Bei, Y. (2020). Medical health big data classification based on KNN classification algorithm. IEEE Access.

Zhang, S. (2022). Challenges in KNN classification. IEEE Transactions on Knowledge and Data Engineering, 34(10), 4663–4675.

Zhang, X., Li, M., Zong, X., Zhu, X., & Wang, R. (2018). Efficient kNN classification with different numbers of nearest neighbors. IEEE Transactions on Neural Networks and Learning Systems.